Metric Spaces with Expensive Distances

### 2020-09-20 paper/draft_ijcga.tex
ijcga draft

\documentclass[a4paper,UKenglish,cleveref, autoref]{socg-lipics-v2019}

\bibliographystyle{plainurl}% the mandatory bibstyle

\input{my_preamble.tex}

\title{Metric Spaces with Expensive Distances}
\titlerunning{Metric Spaces with Expensive Distances}
% uncomment for LIPICS 
%\author{Anonymous Authors}{Anonymous}{}{}{}
\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
\authorrunning{M. Kerber and A. Nigmetov}
%\authorrunning{Anonymous Authors}

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
\ccsdesc[500]{Theory of computation~Random search heuristics}
%% \category{dummy}

% \category{dummy}

\Copyright{Michael Kerber and Arnur Nigmetov}
%\Copyright{Anonymous Authors}
% END uncomment for LIPICS

@@

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see \cref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turning the method useless in the standard computational model.

@@

\myparagraph{Outline}
We review the properties of spanner algorithms in \cref{sec:back}.
In \cref{sec:blind_spanners}, we introduce different
variants of the blind spanner algorithm. Experimental
results on spanners are presented in \cref{sec:experiment_spanners}.
Then we move to the ANN problem, giving
an outline of the proof of the $O(\log n)$ bound in \cref{sec:ann}.
The full proof can be found in the appendix.
We report our experimental results for ANN in \cref{sec:experiment_ann}.

@@

We performed three types of experiments: on points sampled from a low-dimensional Euclidean
space, on images with Earth Mover's distance, and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

\cref{fig:spanner_sparseness}
shows the number of edges of the spanner for various variants of 
blind and non-blind spanner constructions.
Note that for all blind spanner variants, the number of computed distances
is equal to the spanner size, while for the non-blind greedy spanner,
this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
by the size of the spanner.
We can see that, even though none of the blind spanners produces
spanners of the same quality as the standard greedy algorithm,
\bgrdy and all variants of \brndm
perform significantly better than both variants of \textsc{BlindQuasiSorted}.

\ignore{
As for different variants of the \brndm algorithm,
we note that their performance is almost the same, and the algorithm works
significantly better than \textsc{QuasiSorted} variants, but obviously
worse than the blind greedy variant. There is a consistent, though small,
difference between the variants that do not force lower bounds first and
the other two variants of the \brndm.  (see \cref{fig:blind_rbr_variants}). 
}

@@

From our experiments, we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners. More data from experiments in $\RR^n$ can be found in
\cref{sec:spanners_euclidean_addendum}.

@@

Sparseness values of the \bgrdy spanner on the whole \dtype{original} data set
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
the advantage of our algorithm fades away. A plausible explanation
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in size of the greedy spanner.
For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
computations for even $\eps = 0.1$; for a $1.5$-approximation, we
reduce the number of distances computed by a factor of 10.

@@

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with the \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
We compare our approach with the time to compute all pairwise-distance between the
persistence diagrams with \hera, using the same approximation parameter
$\eps$ (that is the reason why the BF column contains different values)\footnote{Let us be completely clear. The time in the BF column is the total running time for computing 
$\wsdist{q}$ with the corresponding $\eps$, so \textit{all} distances in the BF case are approximate. The time in the BG column
is the total time of the \bgrdy algorithm, which internally computes $\wsdist{q}$ with $\eps = 0.01$. In particular,
the sparseness column in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness} shows the fraction of distances that are almost exact.}.
Let us emphasize that the diagrams we have are rather small,
with average cardinality of
350 points. Nevertheless, on this data set the \bgrdy spanner
provides a substantial improvement in terms of real running time,
and we expect this advantage to grow for diagrams of larger cardinality.

In order to see the growth of the spanner size, we use the \dtype{perturbed}
data set, drawing from it random samples.
We plot the sparseness of the spanners obtained by the \grdy
and the \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
for different values of $\eps$, always using the $\wsdist{1}$ distance.
As expected, the ratio of computed distances decreases as the number of diagrams
grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
the growth of the spanner size relative to the number of diagrams.
For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
linear growth) for larger values of $n$.

@@

\myparagraph{Earth Mover's Distance} 
We took the standard approach \cite{rubner2000earth} of transforming a colored image
to the \textit{Lab} color space and creating a $16\times 16\times 16$
distribution histogram. After that we compute the distance between normalized
distributions with \cite{doran_pyemd}.
We used the Caltech 101 dataset \cite{caltech101},
randomly selecting 200 images from different (randomly chosen)
categories.  We measured only the time needed to compute
the Earth Mover's distance, without preprocessing (color space transformation
and binning). The results are presented in \cref{tbl:emd_lab_caltech}.
The first two columns of the table compare the sparseness of the \bgrdy and the greedy spanners.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
Even for the small relative error $\eps = 0.2$,
the \bgrdy spanner saves about $28\%$ of time,
and, if {$2$-approximation} is acceptable,
then the \bgrdy algorithm speeds up
the computation by a factor of almost $5$.

Alternatively, we also experimented with the Earth Mover’s distance on
gray value images by considering each image as a distribution on pixels.
For this (non-standard) variant, we obtain even larger speed-ups~-- see
\cref{sec:spanners_euclidean_addendum} for details.

@@

The full proof is given in \cref{sec:proof_ann},
we summarize the main ideas:
first, we notice that points ``far away'' from the current candidate $c$ 
can be disregarded in the algorithm. More precisely, if $p_i$
is outside of the ball of radius $2\dist(c, q)$ centered at $c$,
the lower bound for $p_i$ will be so large that
the distance to of $p_i$ to $q$ does not have to be queried.
Second, whenever a distance of $q$ to some $p_j$ is computed,
we can disregard all points ``very close'' to $p_j$.
Precisely, points in a ball of radius $\eps\dist(p_j,q)$
will not invalidate the current candidate 
to be an approximate nearest neighbor and hence,
no distance query for such points is needed.

@@

The proof strategy fails for $\eps=0$ because the $\eps$-ball
around $p_j$ as described above degenerates, and the packing argument fails.
Indeed, as the example in \cref{fig:exact_bad_example} (in~\cref{sec:proof_ann}) 
shows, there are point sets where the expected number
of distance computations for exact nearest neighbor is linear.

@@

In order to empirically verify the upper bound $O(\log n)$,
we plot the number of computed distances divided
by the logarithm of the number of points in \cref{fig:ann_const_ratio} (for $d = 2$).
We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
the theoretical upper bound, but also shows that the algorithm in the low-dimensional case
really computes only a very small number of distances to the query point.
As expected, in high dimensions the algorithm does not perform as well.
In \cref{fig:ann_dimension_dependence} we plot the average number of computed distances
for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
the sublinearity of the growth becomes clear only when the number of points is relatively large,
approaching 30000.

Similar experiments were performed on \dtype{perturbed} data set of persistence
diagrams, which was described in \cref{sec:experiment_spanners}.
For a given number $n$, we randomly select $n$ diagrams from the data set
and from the remaining diagrams we randomly choose 20 query points.
In \cref{fig:ann_const_ratio_mcgill} we plot the average ratio of computed
distances to $\log(n)$ for $\wsdist{1}$ and $\wsdist{2}$ metrics.
The average ratio appears to fluctuate around 7 for $\wsdist{1}$ and around 8 for $\wsdist{2}$,
which even makes the algorithm
reasonable for practical application, if pairwise distances are already available.

@@

%As for different variants of the \brndm algorithm,
%we note that their performance is almost the same, and the algorithm works
%significantly better than \textsc{QuasiSorted} variants, but obviously
%worse than the blind greedy variant. There is a consistent, though small,
%difference between the variants that do not force lower bounds first and
%the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }

@@

\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the \bgrdy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \bgrdy and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

@@

The plot in \cref{fig:blind_greedy_dimensions}
compares performance of \bgrdy in different dimensions.
We see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
edges for $700$ points, which clearly shows some degrading for higher dimensions.

The plot in \cref{fig:spanner_eps_dependence}
compares the \bgrdy and \textsc{Greedy} algorithms
on \textbf{uniform} point sets for different choices of $\eps$. Dependence on $\eps$
is approximately the same for both algorithms. Since it is not clearly seen from the picture, we also note
that the ratio of the number of edges decreases for
smaller values of $\eps$: 
for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
while for $\eps = 1/32$ the ratio is 2.6

@@

We randomly selected 250 images from the dataset.
The results of the experiment are summarized in \cref{tbl:emd_gtsrb}.
As in \cref{sec:experiment_spanners}, 
the first two columns give the sparseness.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns compare the running time.
Note that even for $\eps=0.1$,
the \bgrdy spanner saves $70\%$ of the computation time.
If we allow large relative error,
we see the speed up by almost a factor of $10$.

\section{Proof of \cref{thm:ann_bound}}
\label{sec:proof_ann}

@@

\end{lemma}
\begin{proof}
We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
balls of radius $R/4$, etc. Repeating this process $m := \lceil \log \frac{R}{r/2} \rceil$ times, 
we cover
$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
contain at most one point from $P$, 
\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\qedhere\]
\end{proof}

@@

\cref{lem:bound_lemma}
can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
further distance computations of points very close to $p_i$ or very far from $p_i$
will be avoided.

@@

In \cref{fig:ann_illustration} we illustrate this. First, $p_1$ is
chosen as the current candidate, and we must compute $\dist(p_2, q)$.
After that the algorithm will not compute
distance to any of the points inside the heavily shaded ball or outside
the lightly shaded ball that are centered at $p_2$,
because their lower bounds allow us to discard them.

@@

\begin{lemma}
\label{lem:sequence_lemma}
Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
\[\left(\frac{4(2+\eps) v_k}{\eps v_\ell}\right)^{d}\]
distances to $q$.
\end{lemma}
\begin{proof}
By the first part of \cref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
is computed, the second part of \cref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying \cref{lem:packing_lemma}.
See \cref{fig:ann_illustration}.
\end{proof}

A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
the number of computed distances is a constant (since $v_k=v_\ell$). 
This means that to prove \cref{thm:ann_bound}, it would suffice to show
that the candidate changes only a logarithmic number of times in expectation.
Let us consider \cref{fig:ann_illustration} again. Notice that the point $p_5$, which
is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points 
$p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are farther from 
$q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the 
candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, 
thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis,
and we need to modify it slightly.

>>>>

\documentclass{ws-ijcga}

\bibliographystyle{ws-ijcga}

\usepackage{amsmath}%
\usepackage{amssymb}%
\usepackage{amsfonts}
%\usepackage{amsthm}

\usepackage{enumerate}
\usepackage{booktabs}
\usepackage{nccmath}
\usepackage{diagbox}
\usepackage{url}
\usepackage{xspace}
\usepackage{mathbbol}
\usepackage{comment}
\usepackage{braket}
%\usepackage{mathptmx}

% graphics
\usepackage{tikz}
\usepackage{pgf}
\usepackage{graphicx}
\usepackage{pgfplots}
\pgfplotsset{compat=1.14}
\usepackage{pgfplotstable}
\usetikzlibrary{patterns}
\usetikzlibrary{external}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{color}
\usepackage{mwe}
\usepackage{floatflt}

\usetikzlibrary{arrows}

% copied from hera paper - check
\usepackage{subcaption}

%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{proposition}[theorem]{Proposition}
%\newtheorem{corollary}[theorem]{Corollary}
%\newtheorem{definition}[theorem]{Definition}

% copied from expensive distances paper
%
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

% disable ugly latex defaults
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\renewcommand{\epsilon}{\varepsilon}
\renewcommand{\phi}{\varphi}
\newcommand{\eps}{\varepsilon}

% hyphen in math mode
\mathchardef\mhyphen="2D

% common rings/fields/etc
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\ZnZ}{\mathbb{Z}/n\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\HH}{\mathbb{H}}
\newcommand{\Fp}{\mathbb{F}_p}

\newcommand{\hence}{\implies}
\newcommand{\ergo}{\implies}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\rk}{rank}
\DeclareMathOperator*{\im}{im}
\DeclareMathOperator*{\epi}{epi}
% \char command is already defined
\DeclareMathOperator*{\fchar}{char}

%\theoremstyle{definition}
%\newtheorem{mythm}{Theorem}
%\newtheorem{mylem}{Lemma}
%\newtheorem{mycor}{Corollary}
%\newtheorem{myprop}{Proposition}
%\newtheorem{mydef}{Definition}

\newcommand{\yaxis}{$y$\nobreakdash-axis\xspace}
\newcommand{\xaxis}{$x$\nobreakdash-axis\xspace}

\newcommand{\algname}[1]{{\textsc{#1}}\xspace}

% Hera stuff

\newcommand{\ignore}[1]{}

\newcommand{\dtype}[1]{{\textit{\small #1}}}

\newcommand{\libraryname}[1]{{\textsc{#1}}\xspace}

\newcommand{\hera}{\libraryname{Hera}}
\newcommand{\Hera}{\libraryname{Hera}}
\newcommand{\dionisus}{\libraryname{Dionysus}}
\newcommand{\Dionisus}{\libraryname{Dionysus}}

\pgfplotsset
{
    tick label style={font=\footnotesize},
    legend style    ={font=\footnotesize},
    label style     ={font=\footnotesize},
    tickwidth       =2pt,
    ylabel near ticks,
    xlabel near ticks,
    legend style     ={draw=none},
    legend cell align=right,
    legend plot pos  =right,
    /pgfplots/narrow area legend/.style=
    {
        legend image code/.code={ \draw[#1] (0cm,-0.06cm) rectangle (0.6cm,0.06cm); }
    },
    /pgfplots/short line legend/.style=
    {
        legend image code/.code={ \draw[#1] (0cm,0cm) rectangle (0.6cm,0cm); }
    },
}

\pgfplotscreateplotcyclelist{filled-color}
{
    every mark/.append style={solid,fill=blue!80!white},    mark=*          \\%
    every mark/.append style={solid,fill=red!80!white},     mark=square*    \\%
    every mark/.append style={solid,fill=orange!80!white},  mark=pentagon*  \\%
    every mark/.append style={solid,fill=green},            mark=diamond*   \\%
    every mark/.append style={solid,fill=black},            mark=triangle*  \\%
}

\newcommand{\addaveraged}[2][]
           {\addplot+[#1,error bars/.cd, y dir=both,y explicit] table[x index=0, y index=1, y error index=2] {#2};}

\pgfplotsset{
    geommark/.style={black,mark options={solid,fill=red!80!white},mark=square*},
    nongeommark/.style={black,mark options={solid,fill=blue!80!white},mark=*},
    dionmark/.style={black,mark options={solid,fill=orange!80!white},mark=pentagon*},
    greentrianglemark/.style={black,mark options={solid,fill=green!80!white},mark=triangle*},
    browndiamondmark/.style={black,mark options={solid,fill=brown!80!white},mark=diamond*}
}

% all distances
% interleaving
\newcommand{\intdist}{{D_I}}
% bottleneck
\newcommand{\btdist}{{W_\infty}}
% Wasserstein of order q (parameter)
\newcommand{\wsdist}[1]{{W_{#1}}}

% expensive distances stuff
\input{my_preamble_exp_dist.tex}
\title{Metric Spaces with Expensive Distances}

\author{Michael Kerber}

%\address{University Department, University Name, Address\\
%City, State ZIP/Zone,Country\,\footnote{State completely without
%abbreviations, the affiliation and mailing address,
%including country. Typeset in 8 pt italic.}\\
%e-mail address
%}

\author{Arnur Nigmetov}

%\address{Group, Laboratory, Address\\
%City, State ZIP/Zone, Country\\
%e-mail address
%}

@@

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turning the method useless in the standard computational model.

@@

\myparagraph{Outline}
We review the properties of spanner algorithms in Section~\ref{sec:back}.
In Section~\ref{sec:blind_spanners}, we introduce different
variants of the blind spanner algorithm. Experimental
results on spanners are presented in Section~\ref{sec:experiment_spanners}.
Then we move to the ANN problem, giving
an outline of the proof of the $O(\log n)$ bound in Section~\ref{sec:ann}.
The full proof can be found in the appendix.
We report our experimental results for ANN in Section~\ref{sec:experiment_ann}.

@@

We performed three types of experiments: on points sampled from a low-dimensional Euclidean
space, on images with Earth Mover's distance, and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See Section~\ref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

Figure~\ref{fig:spanner_sparseness}
shows the number of edges of the spanner for various variants of 
blind and non-blind spanner constructions.
Note that for all blind spanner variants, the number of computed distances
is equal to the spanner size, while for the non-blind greedy spanner,
this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
by the size of the spanner.
We can see that, even though none of the blind spanners produces
spanners of the same quality as the standard greedy algorithm,
\bgrdy and all variants of \brndm
perform significantly better than both variants of \textsc{BlindQuasiSorted}.

@@

From our experiments, we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners. More data from experiments in $\RR^n$ can be found in
Section~\ref{sec:spanners_euclidean_addendum}.

@@

Sparseness values of the \bgrdy spanner on the whole \dtype{original} data set
for different values of $\eps$ and $q$ are presented in Table~\ref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
the advantage of our algorithm fades away. A plausible explanation
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in size of the greedy spanner.
For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
computations for even $\eps = 0.1$; for a $1.5$-approximation, we
reduce the number of distances computed by a factor of 10.

@@

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with the \bgrdy spanner.
The timings are given in Table~\ref{tbl:mcgill_original_blind_greedy_spanner_timing}.
We compare our approach with the time to compute all pairwise-distance between the
persistence diagrams with \hera, using the same approximation parameter
$\eps$ (that is the reason why the BF column contains different values)\footnote{Let us be completely clear. The time in the BF column is the total running time for computing 
$\wsdist{q}$ with the corresponding $\eps$, so \textit{all} distances in the BF case are approximate. The time in the BG column
is the total time of the \bgrdy algorithm, which internally computes $\wsdist{q}$ with $\eps = 0.01$. In particular,
the sparseness column in Table~\ref{tbl:mcgill_original_blind_greedy_spanner_sparseness} shows the fraction of distances that are almost exact.}.
Let us emphasize that the diagrams we have are rather small,
with average cardinality of
350 points. Nevertheless, on this data set the \bgrdy spanner
provides a substantial improvement in terms of real running time,
and we expect this advantage to grow for diagrams of larger cardinality.

In order to see the growth of the spanner size, we use the \dtype{perturbed}
data set, drawing from it random samples.
We plot the sparseness of the spanners obtained by the \grdy
and the \bgrdy algorithms in Figure~\ref{fig:spanner_sparseness_mcgill_perturbed}
for different values of $\eps$, always using the $\wsdist{1}$ distance.
As expected, the ratio of computed distances decreases as the number of diagrams
grows. In Figure~\ref{fig:spanner_ratio_mcgill_perturbed} we show
the growth of the spanner size relative to the number of diagrams.
For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
linear growth) for larger values of $n$.

@@

\myparagraph{Earth Mover's Distance} 
We took the standard approach \cite{rubner2000earth} of transforming a colored image
to the \textit{Lab} color space and creating a $16\times 16\times 16$
distribution histogram. After that we compute the distance between normalized
distributions with \cite{doran_pyemd}.
We used the Caltech 101 dataset \cite{caltech101},
randomly selecting 200 images from different (randomly chosen)
categories.  We measured only the time needed to compute
the Earth Mover's distance, without preprocessing (color space transformation
and binning). The results are presented in Table~\ref{tbl:emd_lab_caltech}.
The first two columns of the table compare the sparseness of the \bgrdy and the greedy spanners.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
Even for the small relative error $\eps = 0.2$,
the \bgrdy spanner saves about $28\%$ of time,
and, if {$2$-approximation} is acceptable,
then the \bgrdy algorithm speeds up
the computation by a factor of almost $5$.

Alternatively, we also experimented with the Earth Mover’s distance on
gray value images by considering each image as a distribution on pixels.
For this (non-standard) variant, we obtain even larger speed-ups~-- see
Section~\ref{sec:spanners_euclidean_addendum} for details.

@@

The full proof is given in Section~\ref{sec:proof_ann},
we summarize the main ideas:
first, we notice that points ``far away'' from the current candidate $c$ 
can be disregarded in the algorithm. More precisely, if $p_i$
is outside of the ball of radius $2\dist(c, q)$ centered at $c$,
the lower bound for $p_i$ will be so large that
the distance to of $p_i$ to $q$ does not have to be queried.
Second, whenever a distance of $q$ to some $p_j$ is computed,
we can disregard all points ``very close'' to $p_j$.
Precisely, points in a ball of radius $\eps\dist(p_j,q)$
will not invalidate the current candidate 
to be an approximate nearest neighbor and hence,
no distance query for such points is needed.

@@

The proof strategy fails for $\eps=0$ because the $\eps$-ball
around $p_j$ as described above degenerates, and the packing argument fails.
Indeed, as the example in Figure~\ref{fig:exact_bad_example} (in Section~\ref{sec:proof_ann}) 
shows, there are point sets where the expected number
of distance computations for exact nearest neighbor is linear.

@@

In order to empirically verify the upper bound $O(\log n)$,
we plot the number of computed distances divided
by the logarithm of the number of points in Figure~\ref{fig:ann_const_ratio} (for $d = 2$).
We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
the theoretical upper bound, but also shows that the algorithm in the low-dimensional case
really computes only a very small number of distances to the query point.
As expected, in high dimensions the algorithm does not perform as well.
In Figure~\ref{fig:ann_dimension_dependence} we plot the average number of computed distances
for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
the sublinearity of the growth becomes clear only when the number of points is relatively large,
approaching 30000.

Similar experiments were performed on \dtype{perturbed} data set of persistence
diagrams, which was described in Section~\ref{sec:experiment_spanners}.
For a given number $n$, we randomly select $n$ diagrams from the data set
and from the remaining diagrams we randomly choose 20 query points.
In Figure~\ref{fig:ann_const_ratio_mcgill} we plot the average ratio of computed
distances to $\log(n)$ for $\wsdist{1}$ and $\wsdist{2}$ metrics.
The average ratio appears to fluctuate around 7 for $\wsdist{1}$ and around 8 for $\wsdist{2}$,
which even makes the algorithm
reasonable for practical application, if pairwise distances are already available.

@@

@@

Figure~\ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the \bgrdy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in Table~\ref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \bgrdy and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

@@

The plot in Figure~\ref{fig:blind_greedy_dimensions}
compares performance of \bgrdy in different dimensions.
We see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
edges for $700$ points, which clearly shows some degrading for higher dimensions.

The plot in Figure~\ref{fig:spanner_eps_dependence}
compares the \bgrdy and \textsc{Greedy} algorithms
on \textbf{uniform} point sets for different choices of $\eps$. Dependence on $\eps$
is approximately the same for both algorithms. Since it is not clearly seen from the picture, we also note
that the ratio of the number of edges decreases for
smaller values of $\eps$: 
for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
while for $\eps = 1/32$ the ratio is 2.6

@@

We randomly selected 250 images from the dataset.
The results of the experiment are summarized in Table~\ref{tbl:emd_gtsrb}.
As in Section~\ref{sec:experiment_spanners}, 
the first two columns give the sparseness.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns compare the running time.
Note that even for $\eps=0.1$,
the \bgrdy spanner saves $70\%$ of the computation time.
If we allow large relative error,
we see the speed up by almost a factor of $10$.

\section{Proof of Theorem~\ref{thm:ann_bound}}
\label{sec:proof_ann}

@@

\end{lemma}
\begin{proof}
We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
balls of radius $R/4$, etc. Repeating this process $m := \lceil \log \frac{R}{r/2} \rceil$ times, 
we cover
$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
contain at most one point from $P$, 
\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\square\]
\end{proof}

@@

Lemma~\ref{lem:bound_lemma}
can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
further distance computations of points very close to $p_i$ or very far from $p_i$
will be avoided.

@@

In Figure~\ref{fig:ann_illustration} we illustrate this. First, $p_1$ is
chosen as the current candidate, and we must compute $\dist(p_2, q)$.
After that the algorithm will not compute
distance to any of the points inside the heavily shaded ball or outside
the lightly shaded ball that are centered at $p_2$,
because their lower bounds allow us to discard them.

@@

\begin{lemma}
\label{lem:sequence_lemma}
Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
\[\left(\frac{4(2+\eps) v_k}{\eps v_\ell}\right)^{d}\]
distances to $q$.
\end{lemma}
\begin{proof}
By the first part of Lemma~\ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
is computed, the second part of Lemma~\ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma~\ref{lem:packing_lemma}.
See Figure~\ref{fig:ann_illustration}.
\end{proof}

A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
the number of computed distances is a constant (since $v_k=v_\ell$). 
This means that to prove Theorem~\ref{thm:ann_bound}, it would suffice to show
that the candidate changes only a logarithmic number of times in expectation.
Let us consider Figure~\ref{fig:ann_illustration} again. Notice that the point $p_5$, which
is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points 
$p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are farther from 
$q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the 
candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, 
thus Theorem~\ref{thm:ann_bound} does not immediately follow from standard backwards analysis,
and we need to modify it slightly.
